\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-06 paper.tex
Updates from Overleaf

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{geometry}%
\usepackage{color}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{nameref}
\usepackage[right]{lineno}
\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{math,calc,positioning}
\usepackage{url}
% JK: turning this off for the moment as I keep clicking through on links
% to the bibliography while reading the text and it's intensely annoying.
% Can reinstate when we're ready to preprint
% \usepackage[hidelinks]{hyperref}

@@

\linenumbers
\title{A general and efficient representation of Ancestral Recombination Graphs}

% First authors
\author[1]{Yan Wong}

% Second authors
\author[2,$\star$]{Anastasia Ignatieva}
\author[3,4,$\star$]{Jere Koskela}

% Middle Authors
\author[5]{Gregor Gorjanc}
\author[6,7]{Anthony W. Wohns}

% Corresponding
\author[1,$\dagger$]{Jerome Kelleher}

\affil[1]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, OX3 7LF, UK}
\affil[2]{Department of Statistics, University of Oxford, OX1 3LB, UK}
\affil[3]{School of Mathematics, Statistics and Physics, Newcastle University, NE1 7RU, UK}
\affil[4]{Department of Statistics, University of Warwick, CV4 7AL, UK}
\affil[5]{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, EH25 9RG, UK}
\affil[6]{Broad Institute of MIT and Harvard, Cambridge, MA 02142, USA}
\affil[7]{Department of Genetics, Stanford University School of Medicine, Stanford, CA 94305, USA}

\affil[$\star$]{Denotes shared second authorship, listed alphabetically}
\affil[$\dagger$]{Denotes corresponding author}

@@

% FIXME this is weak, but gives the basic outline
\begin{abstract}
New developments have made it possible to infer genetic genealogies in
the presence of recombination at scale, enabling many
downstream applications in population and statistical genetics.
The structure representing such recombinant genetic ancestry
is usually referred to as an Ancestral Recombination Graph (ARG),
although there is some confusion about the interpretation and little
agreement on the concrete details.
We propose a concrete definition of ARGs
in terms of genomes and their intervals of genetic inheritance (gARG)
and contrast this with the classical event-based definition (eARG).
The eARG definition is based on the coalescent with recombination
stochastic process, and we show how the assumptions of that process
place limitations on the patterns of inheritance that can be
represented, and that these assumptions are
% Better wording
substantially broken
in modern large-scale datasets.
Representing genetic ancestry
with an eARG also requires complete precision about the ordering
of all recombination events, which we will rarely have sufficient
information to infer.
We show, in contrast, that the gARG definition
can fully capture the richness of modern large-scale datasets,
enables fine-grained
levels of precision about recombination to be represented,
and forms the basis of an efficient computational framework.
\end{abstract}

@@

\section*{Discussion}
Recent breakthroughs have made ARG inference methods practical
at scale for the first time, and there has been a surge of interest
in inference methods, their evaluation and application.
The prospect of ARGs being used routinely in population
and statistical genetics applications is tantalising,
but in reality there is substantial work to be done to
enable this.
A necessary first step is a degree of terminological clarity.
As discussed in Appendix XXX, the term Ancestral Recombination
Graph has several
subtly different interpretations depending on context.
The trend to decouple ARGs from their origins as a stochastic
process and instead use the term as a more general representation of any
recombinant genetic ancestry seems most useful, which we have
tried to clarify and systematise here. Thus
we can think of an ARG as any structure that encodes the
reticulate genetic ancestry of a sample of colinear sequences under
the influence of recombination. The ``genome'' ARG encoding
made explicit here is one way we can concretely
define such recombinant ancestry, which we have shown is both
flexible and efficient.
The flexibility of the gARG encoding contrasts with the classical
``event'' ARG (eARG), which is more limited in what can be described, because of its origins in an idealised mathematical model.

@@

Given that there is no single model that can capture the complexities
of modern datasets across species (and taxa; see discussion of SARS-CoV-2
below), we cannot rely on sampling from a distribution to capture
the uncertainty in our ARG estimates. Indeed, even if the coalescent
was a suitable model for datasets such as UKB and we had a means of
sampling from its distribution of ARGs at the scale of hundreds of thousands
of samples, we could only explore the tiniest corner of such an incompehensibly
large space of possibilities.
Therefore, it is important to systematically
describe and utilise uncertainty about ARG inference.
One approach, enabled by the gARG encoding described here,
is to use polytomies and stacked recombinations as direct topological
descriptions of uncertainty. Other methods to capture, for example,
uncertainty about node ages and breakpoint locations, is an important
aspect of future work.

>>>>

\documentclass{article}
\usepackage[letterpaper,top=2cm,bottom=2.5cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}
\usepackage[round]{natbib}
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{geometry}%
\usepackage{color}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{nameref}
\usepackage[right]{lineno}
\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{math,calc,positioning}
\usepackage{url}
\usepackage[symbol]{footmisc}
% JK: turning this off for the moment as I keep clicking through on links
% to the bibliography while reading the text and it's intensely annoying.
% Can reinstate when we're ready to preprint
% \usepackage[hidelinks]{hyperref}

@@

\linenumbers
\title{\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs}

% % First authors
% \author[1]{Yan Wong}

% % Second authors
% \author[2,$\star$]{Anastasia Ignatieva}
% \author[3,4,$\star$]{Jere Koskela}

% % Middle Authors
% \author[5]{Gregor Gorjanc}
% \author[6,7]{Anthony W. Wohns}

% % Corresponding
% \author[1,$\dagger$]{Jerome Kelleher}

% \affil[1]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, OX3 7LF, UK}
% \affil[2]{Department of Statistics, University of Oxford, OX1 3LB, UK}
% \affil[3]{School of Mathematics, Statistics and Physics, Newcastle University, NE1 7RU, UK}
% \affil[4]{Department of Statistics, University of Warwick, CV4 7AL, UK}
% \affil[5]{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, EH25 9RG, UK}
% \affil[6]{Broad Institute of MIT and Harvard, Cambridge, MA 02142, USA}
% \affil[7]{Department of Genetics, Stanford University School of Medicine, Stanford, CA 94305, USA}

% \affil[$\star$]{Denotes shared second authorship, listed alphabetically}
% \affil[$\dagger$]{Denotes corresponding author}

\author[1]{Yan~Wong}
\author[2,3$\star$]{Anastasia~Ignatieva}
\author[4,5$\star$]{Jere~Koskela}
\author[6]{Gregor~Gorjanc}
\author[7,8]{Anthony~W.~Wohns}
\author[1$\dagger$]{Jerome~Kelleher}
\affil[ ]{\mbox{}\vspace{-2.5em}}

@@

\setlength{\skip\footins}{1em}
\setlength{\footnotemargin}{0.5em}
\renewcommand{\thefootnote}{\arabic{footnote}}
\footnotetext[1]{Big Data Institute, Li Ka Shing Centre for Health
Information and Discovery, University of Oxford, UK}
\footnotetext[2]{School of Mathematics and Statistics, University of Glasgow, UK}
\footnotetext[3]{Department of Statistics, University of Oxford, UK}
\footnotetext[4]{School of Mathematics, Statistics and Physics, Newcastle University, UK}
\footnotetext[5]{Department of Statistics, University of Warwick, UK}
\footnotetext[6]{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, UK}
\footnotetext[7]{Broad Institute of MIT and Harvard, Cambridge, USA}
\footnotetext[8]{Department of Genetics, Stanford University School of Medicine, Stanford, USA}
\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\footnotetext[1]{Shared second authorship, listed alphabetically}
\footnotetext[2]{Correspondence: jerome.kelleher@bdi.ox.ac.uk}

\vspace{-1em}

% FIXME this is weak, but gives the basic outline
\begin{abstract}
New developments have made it possible to infer genetic genealogies in
the presence of recombination at scale, enabling many
downstream applications in population and statistical genetics.
The structure representing such recombinant genetic ancestry
is usually referred to as an ancestral recombination graph (ARG),
although there is some confusion about the interpretation and little
agreement on the concrete details.
We propose a concrete definition of ARGs
in terms of genomes and their intervals of genetic inheritance (gARG)
and contrast this with the classical event-based definition (eARG).
The eARG definition is based on the coalescent with recombination
stochastic process, and we show how the assumptions of that process
place limitations on the patterns of inheritance that can be
represented, and that these assumptions are
% Better wording
substantially broken
in modern large-scale datasets.
Representing genetic ancestry
with an eARG also requires complete precision about the ordering
of all recombination events, which we will rarely have sufficient
information to infer.
We show, in contrast, that the gARG definition
can fully capture the richness of modern large-scale datasets,
enables fine-grained
levels of precision about recombination to be represented,
and forms the basis of an efficient computational framework.
\end{abstract}

@@

\section*{Discussion}
Recent breakthroughs have finally made large-scale ARG inference
feasible in practice, leading to a surge of interest
in development of ARG-based inference methods, their evaluation 
and application.
The prospect of ARGs being used routinely within population
and statistical genetics is tantalising,
but in reality there is substantial work to be done to
enable this.
A necessary first step is a degree of terminological clarity.
As discussed in Appendix XXX, the term ``ancestral recombination
graph" has several
subtly different interpretations, depending on context.
The trend to decouple ARGs from their original definiton 
as a stochastic
process and instead use the term as a more general representation of any
recombinant genetic ancestry seems most useful, which we have
tried to clarify and systematise here. Thus
we can think of an ARG as any structure that encodes the
reticulate genetic ancestry of a sample of colinear sequences under
the influence of recombination. The ``genome'' ARG encoding
made explicit here is one way we can concretely
define such recombinant ancestry, which we have shown is both
flexible and efficient.
The flexibility of the gARG encoding contrasts with the classical
``event'' ARG (eARG), which is more limited in what can be described, because of its origins in an idealised mathematical model.

@@

Given that there is no single model that can capture the complexities
of modern datasets across species (and taxa; see discussion of SARS-CoV-2
below), we cannot rely on sampling from a distribution to capture
the uncertainty in our ARG estimates. Indeed, even if the coalescent
was a suitable model for datasets such as UKB and we had a means of
sampling from its distribution of ARGs at the scale of hundreds of thousands
of samples, we could only explore the tiniest corner of such an incomprehensibly
large space of possibilities.
Therefore, it is important to systematically
describe and utilise uncertainty about ARG inference.
One approach, enabled by the gARG encoding described here, is to allow
nodes to have (rootward) in-degree greater than two (polytomies
representing uncertainty over the ordering of coalescence events) or 
out-degree greater than two (representing uncertainty over the ordering
of multiple recombination events). Other methods to capture, for example,
uncertainty about node ages and breakpoint locations, is an important
aspect of future work.
